quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-10-15 paper-quollr.Rmd
Updated Usage section

The following demonstration of the package's functionality assumes `quollr` has been loaded. The built-in data sets `scurve` and `scurve_umap` are also used throughout.

The `scurve` is a $7\text{-}D$ simulated dataset. It is constructed by simulating $1000$ observations.

`scurve_umap` is the UMAP \twoD embedding for `scurve` data with `n_neighbors`$=46$ and `min_dist`$=0.9$. Each data set contains a unique ID column that maps the $7\text{-}D$ point in `scurve` to the corresponding \twoD `scurve_umap`.

>>>>

The following demonstration of the package's functionality assumes `quollr` has been loaded. To begin the workflow, users need two inputs: the high-dimensional dataset and the corresponding nonlinear dimensionality reduction (NLDR) layout. The high-dimensional data must contain a unique `ID` column, with data columns prefixed by the letter `"x"` (e.g., `x1`, `x2`, etc.). The NLDR dataset should include embedding coordinates labeled as `emb1` and `emb2`, ensuring one-to-one correspondence with the high-dimensional data through the shared `ID`.

To illustrate the workflow, we use the built-in example dataset `scurve`, a $7\text{-}D$ simulated dataset consisting of $1000$ observations. The first three variables define a $3\text{-}D$ S-shaped manifold, while the remaining four variables introduce low-magnitude uniform noise, yielding a structured yet noisy high-dimensional dataset.

```{r, echo=TRUE}
data("scurve")
```

For this example, we use the UMAP layout, which is produced using `n_neighbors =` $46$ and `min_dist =` $0.9$.

```{r, echo=TRUE}
library(umap)
library(dplyr)

scurve_umap <- umap(
  scurve |> select(-ID),
  config = modifyList(umap.defaults, list(
    n_neighbors = 46,
    n_components = 2,
    min_dist = 0.9
  ))
)$layout |>
  as_tibble(.name_repair = ~ paste0("emb", 1:2)) |>
  mutate(ID = scurve$ID)
```

### 2025-10-15 paper-quollr.tex
Updated Usage section

The following demonstration of the package's functionality assumes \texttt{quollr} has been loaded. The built-in data sets \texttt{scurve} and \texttt{scurve\_umap} are also used throughout.

The \texttt{scurve} is a \(7\text{-}D\) simulated dataset. It is constructed by simulating \(1000\) observations.

\texttt{scurve\_umap} is the UMAP $2\text{-}D$ embedding for \texttt{scurve} data with \texttt{n\_neighbors}\(=46\) and \texttt{min\_dist}\(=0.9\). Each data set contains a unique ID column that maps the \(7\text{-}D\) point in \texttt{scurve} to the corresponding $2\text{-}D$ \texttt{scurve\_umap}.

@@

\begin{verbatim}
> $scaled_nldr
> # A tibble: 1,000 x 3
>     emb1  emb2    ID
>    <dbl> <dbl> <int>
>  1 0.277 0.913     1
>  2 0.697 0.538     2
>  3 0.779 0.399     3
>  4 0.173 0.953     4
>  5 0.218 0.983     5
>  6 0.593 1.05      6
>  7 0.180 0.210     7
>  8 0.976 0.571     8
>  9 0.803 0.829     9
> 10 0.932 0.410    10
> # i 990 more rows
> 
> $lim1
> [1] -9.146398  8.552211
> 
> $lim2
> [1] -10.36686  10.10691
\end{verbatim}

@@

\begin{verbatim}
> $b2
> [1] 33
> 
> $a1
> [1] 0.04952516
> 
> $a2
> [1] 0.04289005
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 825 x 3
>        h       c_x    c_y
>    <int>     <dbl>  <dbl>
>  1     1 -0.1      -0.116
>  2     2 -0.0505   -0.116
>  3     3 -0.000950 -0.116
>  4     4  0.0486   -0.116
>  5     5  0.0981   -0.116
>  6     6  0.148    -0.116
>  7     7  0.197    -0.116
>  8     8  0.247    -0.116
>  9     9  0.296    -0.116
> 10    10  0.346    -0.116
> # i 815 more rows
\end{verbatim}

@@

\begin{verbatim}
>   h           x           y
> 1 1 -0.10000000 -0.08708678
> 2 1 -0.12476258 -0.10138346
> 3 1 -0.12476258 -0.12997683
> 4 1 -0.10000000 -0.14427351
> 5 1 -0.07523742 -0.12997683
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 4
>    emb1  emb2    ID     h
>   <dbl> <dbl> <int> <int>
> 1 0.277 0.913     1   609
> 2 0.697 0.538     2   392
> 3 0.779 0.399     3   319
> 4 0.173 0.953     4   631
> 5 0.218 0.983     5   657
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 3
>       h   n_h   w_h
>   <int> <int> <dbl>
> 1    82     1 0.001
> 2    83     4 0.004
> 3    84     1 0.001
> 4    85     3 0.003
> 5    86     1 0.001
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 2
>       h pts_list 
>   <int> <list>   
> 1    82 <int [1]>
> 2    83 <int [4]>
> 3    84 <int [1]>
> 4    85 <int [3]>
> 5    86 <int [1]>
\end{verbatim}

@@

\begin{verbatim}
>   h           c_x        c_y n_h w_h
> 1 1 -0.1000000000 -0.1156801   0   0
> 2 2 -0.0504748374 -0.1156801   0   0
> 3 3 -0.0009496749 -0.1156801   0   0
> 4 4  0.0485754877 -0.1156801   0   0
> 5 5  0.0981006503 -0.1156801   0   0
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 8
>    from    to    x_from  y_from    x_to    y_to from_count to_count
>   <int> <int>     <dbl>   <dbl>   <dbl>   <dbl>      <dbl>    <dbl>
> 1     1     2 -0.1      -0.116  -0.0505 -0.116           0        0
> 2    26    27 -0.0752   -0.0728 -0.0257 -0.0728          0        0
> 3     3    27 -0.000950 -0.116  -0.0257 -0.0728          0        0
> 4    26    52 -0.0752   -0.0728 -0.0505 -0.0299          0        0
> 5    27    28 -0.0257   -0.0728  0.0238 -0.0728          0        0
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 10
>    from    to    x_from  y_from    x_to    y_to from_count to_count
>   <int> <int>     <dbl>   <dbl>   <dbl>   <dbl>      <dbl>    <dbl>
> 1     1     2 -0.1      -0.116  -0.0505 -0.116           0        0
> 2    26    27 -0.0752   -0.0728 -0.0257 -0.0728          0        0
> 3     3    27 -0.000950 -0.116  -0.0257 -0.0728          0        0
> 4    26    52 -0.0752   -0.0728 -0.0505 -0.0299          0        0
> 5    27    28 -0.0257   -0.0728  0.0238 -0.0728          0        0
> # i 2 more variables: from_reindexed <int>, to_reindexed <int>
\end{verbatim}

@@

\begin{verbatim}
>   [1]  82  83  84  85  86  93  94 107 108 109 110 111 112 113 118 119 120 131
>  [19] 132 133 134 135 136 137 138 141 142 143 144 145 156 157 158 159 160 161
>  [37] 162 163 164 165 166 167 168 169 170 171 181 182 183 184 185 186 187 188
>  [55] 189 190 191 192 193 194 195 196 206 207 209 210 211 212 213 214 215 216
>  [73] 217 218 219 220 221 222 231 232 233 234 235 236 237 239 240 241 242 243
>  [91] 244 245 246 256 257 258 259 260 261 266 267 268 269 271 272 281 282 283
> [109] 284 285 291 292 293 294 295 296 297 307 317 318 319 320 321 322 323 341
> [127] 342 343 344 345 346 347 367 368 369 371 372 373 391 392 393 394 395 396
> [145] 397 398 416 417 418 419 420 421 422 423 440 441 442 443 444 446 447 465
> [163] 466 467 468 469 470 471 472 489 490 491 492 493 494 495 496 497 509 510
> [181] 511 512 513 514 515 516 517 518 519 520 521 522 528 530 531 532 533 534
> [199] 535 536 537 538 539 540 541 542 543 544 545 546 553 554 555 556 557 558
> [217] 559 560 561 562 563 564 565 566 567 568 569 570 578 579 580 581 582 583
> [235] 584 585 586 587 588 589 590 591 592 593 594 595 603 604 605 606 607 608
> [253] 609 610 611 612 613 614 615 616 617 618 619 620 628 629 630 631 633 634
> [271] 635 636 637 638 640 641 642 643 644 653 654 655 656 657 658 659 660 661
> [289] 662 663 664 665 666 667 678 679 680 681 682 683 685 686 687 688 689 690
> [307] 691 704 705 706 707 708 709 710 711 712 713 714 729 730 731 732 733 734
> [325] 735 736 737 738 755 757
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 8
>       h    x1     x2    x3       x4       x5      x6       x7
>   <int> <dbl>  <dbl> <dbl>    <dbl>    <dbl>   <dbl>    <dbl>
> 1    82 0.888 0.102   1.46  0.0145  -0.0118   0.0303 -0.00707
> 2    83 0.744 0.123   1.66 -0.00950  0.00499 -0.0208 -0.00425
> 3    84 0.591 0.0447  1.81 -0.0168   0.0173  -0.0485 -0.00112
> 4    85 0.371 0.0937  1.93 -0.00109  0.00840  0.0268 -0.00286
> 5    86 0.255 0.0447  1.97 -0.00862 -0.00733 -0.0826  0.00229
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 4
>   pred_emb_1 pred_emb_2    ID pred_h
>        <dbl>      <dbl> <int>  <int>
> 1      0.296      0.914     1    609
> 2      0.717      0.528     2    392
> 3      0.791      0.399     3    319
> 4      0.172      0.957     4    631
> 5      0.197      0.999     5    657
\end{verbatim}

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/unnamed-chunk-27-1} \caption{$2\text{-}D$ projections of the lifted high-dimensional wireframe model from the `Scurve` UMAP layout. Each panel (a1–a4) shows the model (black) overlaid on `Scurve` data (in purple) in different projections. These views illustrate how the lifted wireframe model captures the structure of the `Scurve` data. The two twists visible in the UMAP layout can also be seen in the lifted model.}\label{fig:unnamed-chunk-27}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/unnamed-chunk-30-1} \caption{Exploring the correspondence between UMAP layout and `Scurve` structure in $7\text{-}D$. Two sets of plots are linked: UMAP layout (a1, b1) and projection of $7\text{-}D$ model and data (a2, b2). The purple points indicate the selected subsets, which differ between rows. In (a1), the lower bridge of the `Scurve` is highlighted, which corresponds in (a2) to points spanning across both arms of the high-dimensional structure. In (b1), a different region near the upper arm of the `Scurve` is selected, and in (b2) these points map onto one side of the curved manifold in $7\text{-}D$ projection. While the UMAP layout suggests distinct local clusters, the linked tour views reveal how these selections trace continuous structures in the $7\text{-}D$ space, highlighting distortions introduced by UMAP.}\label{fig:unnamed-chunk-30}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/unnamed-chunk-34-1} \caption{Exploring residuals in relation to UMAP layouts using a $7\text{-}D$ `Scurve` model. Three views are linked: distribution of residuals (a1, b1), UMAP layout (a2, b2), and projection of the $7\text{-}D$ model with data (a3, b3). The purple points highlight selected subsets of the data, which differ across rows. In the top row (a1–a3), points with higher residuals (a1) are selected, corresponding to the sparse bridging region in the UMAP layout (a2) and the less dense end of the `Scurve` in the high-dimensional projection (a3). In the bottom row (b1–b3), points with lower residuals (b1) are highlighted, which map to one side of the dense region in the NLDR layout (b2) and to a thicker band of the `Scurve` in the projection (b3). This comparison illustrates how residuals can help diagnose distortions in UMAP, with high-residual points often concentrated in sparse or stretched regions of the structure.}\label{fig:unnamed-chunk-34}
\end{figure}

>>>>

The following demonstration of the package's functionality assumes \texttt{quollr} has been loaded. To begin the workflow, users need two inputs: the high-dimensional dataset and the corresponding nonlinear dimensionality reduction (NLDR) layout. The high-dimensional data must contain a unique \texttt{ID} column, with data columns prefixed by the letter \texttt{"x"} (e.g., \texttt{x1}, \texttt{x2}, etc.). The NLDR dataset should include embedding coordinates labeled as \texttt{emb1} and \texttt{emb2}, ensuring one-to-one correspondence with the high-dimensional data through the shared \texttt{ID}.

To illustrate the workflow, we use the built-in example dataset \texttt{scurve}, a \(7\text{-}D\) simulated dataset consisting of \(1000\) observations. The first three variables define a \(3\text{-}D\) S-shaped manifold, while the remaining four variables introduce low-magnitude uniform noise, yielding a structured yet noisy high-dimensional dataset.

\begin{verbatim}
data("scurve")
\end{verbatim}

For this example, we use the UMAP layout, which is produced using \texttt{n\_neighbors\ =} \(46\) and \texttt{min\_dist\ =} \(0.9\).

\begin{verbatim}
library(umap)
library(dplyr)

scurve_umap <- umap(
  scurve |> select(-ID),
  config = modifyList(umap.defaults, list(
    n_neighbors = 46,
    n_components = 2,
    min_dist = 0.9
  ))
)$layout |>
  as_tibble(.name_repair = ~ paste0("emb", 1:2)) |>
  mutate(ID = scurve$ID)
\end{verbatim}

@@

\begin{verbatim}
> $scaled_nldr
> # A tibble: 1,000 x 3
>     emb1  emb2    ID
>    <dbl> <dbl> <int>
>  1 0.300 0.985     1
>  2 0.721 0.688     2
>  3 0.722 0.542     3
>  4 0.196 1.01      4
>  5 0.240 1.06      5
>  6 0.621 1.19      6
>  7 0.159 0.160     7
>  8 0.979 0.661     8
>  9 0.870 0.963     9
> 10 0.885 0.501    10
> # i 990 more rows
> 
> $lim1
> [1] -8.320406  8.522628
> 
> $lim2
> [1] -11.732897   9.679631
\end{verbatim}

@@

\begin{verbatim}
> $b2
> [1] 36
> 
> $a1
> [1] 0.04943533
> 
> $a2
> [1] 0.04281225
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 900 x 3
>        h      c_x    c_y
>    <int>    <dbl>  <dbl>
>  1     1 -0.1     -0.127
>  2     2 -0.0506  -0.127
>  3     3 -0.00113 -0.127
>  4     4  0.0483  -0.127
>  5     5  0.0977  -0.127
>  6     6  0.147   -0.127
>  7     7  0.197   -0.127
>  8     8  0.246   -0.127
>  9     9  0.295   -0.127
> 10    10  0.345   -0.127
> # i 890 more rows
\end{verbatim}

@@

\begin{verbatim}
>   h           x           y
> 1 1 -0.10000000 -0.09858838
> 2 1 -0.12471766 -0.11285913
> 3 1 -0.12471766 -0.14140063
> 4 1 -0.10000000 -0.15567138
> 5 1 -0.07528234 -0.14140063
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 4
>    emb1  emb2    ID     h
>   <dbl> <dbl> <int> <int>
> 1 0.300 0.985     1   659
> 2 0.721 0.688     2   492
> 3 0.722 0.542     3   418
> 4 0.196 1.01      4   657
> 5 0.240 1.06      5   708
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 3
>       h   n_h   w_h
>   <int> <int> <dbl>
> 1    82     2 0.002
> 2    83     1 0.001
> 3   106     1 0.001
> 4   107     5 0.005
> 5   108     7 0.007
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 2
>       h pts_list 
>   <int> <list>   
> 1    82 <int [2]>
> 2    83 <int [1]>
> 3   106 <int [1]>
> 4   107 <int [5]>
> 5   108 <int [7]>
\end{verbatim}

@@

\begin{verbatim}
>   h          c_x        c_y n_h w_h
> 1 1 -0.100000000 -0.1271299   0   0
> 2 2 -0.050564674 -0.1271299   0   0
> 3 3 -0.001129349 -0.1271299   0   0
> 4 4  0.048305977 -0.1271299   0   0
> 5 5  0.097741302 -0.1271299   0   0
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 8
>    from    to   x_from  y_from     x_to    y_to from_count to_count
>   <int> <int>    <dbl>   <dbl>    <dbl>   <dbl>      <dbl>    <dbl>
> 1     1     2 -0.1     -0.127  -0.0506  -0.127           0        0
> 2    26    27 -0.0753  -0.0843 -0.0258  -0.0843          0        0
> 3     3    27 -0.00113 -0.127  -0.0258  -0.0843          0        0
> 4    26    52 -0.0753  -0.0843 -0.0506  -0.0415          0        0
> 5    52    53 -0.0506  -0.0415 -0.00113 -0.0415          0        0
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 10
>    from    to   x_from  y_from     x_to    y_to from_count to_count
>   <int> <int>    <dbl>   <dbl>    <dbl>   <dbl>      <dbl>    <dbl>
> 1     1     2 -0.1     -0.127  -0.0506  -0.127           0        0
> 2    26    27 -0.0753  -0.0843 -0.0258  -0.0843          0        0
> 3     3    27 -0.00113 -0.127  -0.0258  -0.0843          0        0
> 4    26    52 -0.0753  -0.0843 -0.0506  -0.0415          0        0
> 5    52    53 -0.0506  -0.0415 -0.00113 -0.0415          0        0
> # i 2 more variables: from_reindexed <int>, to_reindexed <int>
\end{verbatim}

@@

\begin{verbatim}
>   [1]  82  83 106 107 108 109 131 132 133 134 135 156 157 158 159 160 161 162
>  [19] 179 180 181 183 184 185 186 188 189 190 204 205 206 207 208 209 210 211
>  [37] 214 215 216 217 228 229 230 231 232 233 234 235 236 238 239 240 241 242
>  [55] 254 255 256 257 258 259 260 261 263 264 265 266 267 268 269 279 280 281
>  [73] 283 284 285 286 287 288 289 290 293 294 305 306 307 308 309 310 311 312
>  [91] 313 314 315 316 317 318 319 320 321 331 332 333 334 335 336 337 339 340
> [109] 341 342 344 345 346 358 359 360 361 362 363 364 365 366 367 369 370 371
> [127] 372 386 387 388 389 390 391 392 393 394 395 396 414 415 416 417 418 419
> [145] 420 421 422 440 441 442 443 444 445 446 447 467 468 469 470 471 472 473
> [163] 491 492 493 494 495 496 497 498 518 519 520 522 523 541 542 543 544 545
> [181] 546 547 548 554 558 560 566 567 568 569 570 571 572 573 578 579 580 581
> [199] 582 583 584 585 586 587 588 590 591 592 593 594 595 596 597 598 603 604
> [217] 605 606 607 608 609 610 611 612 613 614 615 616 617 618 619 620 621 622
> [235] 623 629 630 631 633 634 635 636 637 638 639 640 641 642 643 644 645 646
> [253] 647 653 654 655 656 657 658 659 660 661 662 663 664 665 666 667 668 669
> [271] 670 671 672 678 679 680 681 682 683 684 685 686 687 688 689 690 691 692
> [289] 693 694 695 696 703 704 705 706 707 708 709 710 711 712 713 714 716 717
> [307] 718 719 720 721 728 729 730 731 732 733 734 736 737 738 739 741 742 743
> [325] 744 754 755 756 757 758 759 760 761 762 763 764 765 766 767 768 779 781
> [343] 782 783 784 785 786 787 788 789 790 791 792 807 808 809 810 811 812 813
> [361] 814 837
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 8
>       h    x1    x2    x3       x4        x5       x6        x7
>   <int> <dbl> <dbl> <dbl>    <dbl>     <dbl>    <dbl>     <dbl>
> 1    82 0.989  1.78  1.13  0.00470  0.0142   -0.0547  -0.00428 
> 2    83 0.960  1.93  1.28  0.00278 -0.0180   -0.0678  -0.00260 
> 3   106 0.997  1.37  1.07  0.00154  0.00687   0.0289   0.00760 
> 4   107 0.975  1.46  1.20 -0.00497 -0.000595  0.0139   0.000537
> 5   108 0.931  1.64  1.36 -0.00706  0.00487  -0.00705  0.000383
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 4
>   pred_emb_1 pred_emb_2    ID pred_h
>        <dbl>      <dbl> <int>  <int>
> 1      0.295      0.986     1    659
> 2      0.716      0.686     2    492
> 3      0.740      0.558     3    418
> 4      0.197      0.986     4    657
> 5      0.246      1.07      5    708
\end{verbatim}

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/unnamed-chunk-29-1} \caption{$2\text{-}D$ projections of the lifted high-dimensional wireframe model from the `Scurve` UMAP layout. Each panel (a1–a4) shows the model (black) overlaid on `Scurve` data (in purple) in different projections. These views illustrate how the lifted wireframe model captures the structure of the `Scurve` data. The two twists visible in the UMAP layout can also be seen in the lifted model.}\label{fig:unnamed-chunk-29}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/unnamed-chunk-32-1} \caption{Exploring the correspondence between UMAP layout and `Scurve` structure in $7\text{-}D$. Two sets of plots are linked: UMAP layout (a1, b1) and projection of $7\text{-}D$ model and data (a2, b2). The purple points indicate the selected subsets, which differ between rows. In (a1), the lower bridge of the `Scurve` is highlighted, which corresponds in (a2) to points spanning across both arms of the high-dimensional structure. In (b1), a different region near the upper arm of the `Scurve` is selected, and in (b2) these points map onto one side of the curved manifold in $7\text{-}D$ projection. While the UMAP layout suggests distinct local clusters, the linked tour views reveal how these selections trace continuous structures in the $7\text{-}D$ space, highlighting distortions introduced by UMAP.}\label{fig:unnamed-chunk-32}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/unnamed-chunk-36-1} \caption{Exploring residuals in relation to UMAP layouts using a $7\text{-}D$ `Scurve` model. Three views are linked: distribution of residuals (a1, b1), UMAP layout (a2, b2), and projection of the $7\text{-}D$ model with data (a3, b3). The purple points highlight selected subsets of the data, which differ across rows. In the top row (a1–a3), points with higher residuals (a1) are selected, corresponding to the sparse bridging region in the UMAP layout (a2) and the less dense end of the `Scurve` in the high-dimensional projection (a3). In the bottom row (b1–b3), points with lower residuals (b1) are highlighted, which map to one side of the dense region in the NLDR layout (b2) and to a thicker band of the `Scurve` in the projection (b3). This comparison illustrates how residuals can help diagnose distortions in UMAP, with high-residual points often concentrated in sparse or stretched regions of the structure.}\label{fig:unnamed-chunk-36}
\end{figure}
